\documentclass[11pt,a4paper]{article}
\input{preamble}

\fancyhead[L]{\small\textsc{tcfs}}
\fancyhead[R]{\small Architecture}

\title{tcfs Architecture}
\author{tcfs / TummyCrypt}
\date{February 2026}

\begin{document}
\maketitle
\tableofcontents
\newpage

% ─────────────────────────────────────────────────────────────────────────────
\section{Overview}

\tcfs{} (TummyCrypt FileSystem) is a FOSS, self-hosted, S3-first synchronization
system that replaces proprietary cloud storage clients (odrive, Dropbox, etc.)
with platform-specific filesystem surfaces that transparently hydrate files on
demand.

\subsection{Design Goals}

\begin{enumerate}
  \item \textbf{Self-hosted first} --- all data stays on user-controlled infrastructure (SeaweedFS).
  \item \textbf{On-demand hydration} --- remote files appear locally as clean
        names or platform placeholders and are downloaded only when accessed.
  \item \textbf{End-to-end encryption} --- client-side XChaCha20-Poly1305 encryption
        before upload; storage operator never sees plaintext.
  \item \textbf{Content-addressed deduplication} --- FastCDC chunking with BLAKE3
        hashing eliminates redundant storage.
  \item \textbf{Cross-platform} --- Linux (FUSE3), macOS/iOS (FileProvider), Windows (Cloud Files API).
  \item \textbf{Kubernetes-native backend} --- stateless sync workers scaled by KEDA,
        NATS JetStream for reliable task dispatch.
\end{enumerate}

% ─────────────────────────────────────────────────────────────────────────────
\section{System Architecture}

The system is divided into client-side and server-side components connected
via S3 API and NATS JetStream.

\subsection{Client Side}

The client runs on user machines and consists of:

\begin{description}
  \item[\tcfsd{} (daemon)] Long-running process exposing a gRPC service over a
    Unix domain socket. Manages FUSE mounts, credential loading, sync state,
    and Prometheus metrics. Reports readiness via \code{sd\_notify(READY=1)}.
  \item[\tcfs{} (CLI)] Command-line interface for push, pull, mount, unmount,
    sync-status, and unsync operations.
  \item[\code{tcfs-tui} (TUI)] Interactive terminal dashboard built with
    ratatui~0.30, connecting to the daemon via gRPC. Five tabs: Dashboard,
    Config, Mounts, Secrets, Conflicts.
  \item[\code{tcfs-mcp} (MCP server)] Model Context Protocol server for AI
    agent integration. Eight tools over stdio JSON-RPC transport.
\end{description}

\subsection{Server Side (Kubernetes)}

\begin{description}
  \item[SeaweedFS cluster] Distributed blob storage. Three Raft masters for
    metadata consensus, volume servers on Drobo 5C for data.
  \item[NATS JetStream] Reliable task queue with two streams:
    \code{SYNC\_TASKS} and \code{HYDRATION\_EVENTS}.
  \item[Sync workers] Stateless \tcfsd{} instances running in
    \code{--mode=worker}. Horizontally scaled by KEDA based on NATS queue depth.
  \item[Prometheus + Grafana] Observability stack monitoring throughput,
    queue depth, and FUSE latency.
\end{description}

% ─────────────────────────────────────────────────────────────────────────────
\section{Crate Map}

All Rust code lives in \filepath{crates/}. The workspace contains 18 product
crates plus an end-to-end test crate:

\begin{longtable}{llp{8cm}}
  \toprule
  \textbf{Crate} & \textbf{Type} & \textbf{Purpose} \\
  \midrule
  \endhead
  \code{tcfs-core}        & library & Shared types, config schema, protobuf defs \\
  \code{tcfs-crypto}      & library & XChaCha20-Poly1305, Argon2id KDF, HKDF, BIP-39 \\
  \code{tcfs-auth}        & library & TOTP, WebAuthn/FIDO2, session, and enrollment helpers \\
  \code{tcfs-secrets}     & library & SOPS/age/KDBX credential chain \\
  \code{tcfs-storage}     & library & OpenDAL abstraction, SeaweedFS native API \\
  \code{tcfs-chunks}      & library & FastCDC chunking, BLAKE3 hashing, zstd compression \\
  \code{tcfs-sync}        & library & Sync engine, JSON state cache, NATS consumers \\
  \code{tcfs-vfs}         & library & Shared virtual filesystem, disk cache, stubs, hydration \\
  \code{tcfs-fuse}        & library & Linux FUSE driver (fuse3 crate) \\
  \code{tcfs-nfs}         & library & NFS loopback mount backend \\
  \code{tcfs-cloudfilter} & library & Windows Cloud Files API (skeleton) \\
  \code{tcfs-sops}        & library & SOPS+age fleet secret propagation \\
  \code{tcfs-file-provider} & library & macOS/iOS FileProvider FFI (RFC 0002) \\
  \code{tcfs-dbus}        & library & Linux D-Bus desktop status integration \\
  \code{tcfsd}            & binary  & Daemon: gRPC, FUSE, metrics, systemd \\
  \code{tcfs-cli}         & binary  & CLI: push, pull, mount, status, unsync \\
  \code{tcfs-tui}         & binary  & TUI: ratatui dashboard \\
  \code{tcfs-mcp}         & binary  & MCP server: 8 tools, stdio transport \\
  \bottomrule
\end{longtable}

% ─────────────────────────────────────────────────────────────────────────────
\section{Physical Stub File Format}

Physical sync-root and compatibility paths can represent files as \code{.tc}
stubs. Mounted VFS views normally expose clean filenames and hydrate from remote
index entries; macOS FileProvider uses platform placeholders / APFS dataless
files. A physical \code{.tc} stub is a small sorted key/value text file:

\begin{lstlisting}
version https://tummycrypt.io/tcfs/v1
chunks 23
compressed 0
fetched 0
oid blake3:4d7a214614ab2935c943f9e0ff69d22eadbb8f32b1258daaa5e2ca24d17e239
origin seaweedfs://filer.example.com/bucket/path/to/file
size 94371840
\end{lstlisting}

Directory stubs use the \code{.tcf} extension for physical directory
dehydration/compatibility paths. Mounted directory listings normally come from
remote \filepath{\{prefix\}/index/...} entries.

% ─────────────────────────────────────────────────────────────────────────────
\section{Hydration Sequence}

When a remote-backed file is opened through the mounted VFS:

\begin{enumerate}
  \item The mount backend's \code{open()} handler intercepts the file access
        request.
  \item \tcfsd{} fetches the manifest from
        \filepath{\{prefix\}/manifests/\{file\_hash\}}.
  \item Chunks are fetched in parallel from
        \filepath{\{prefix\}/chunks/\{chunk\_hash\}}.
  \item Each chunk is decompressed (zstd), decrypted (XChaCha20-Poly1305),
        and verified (BLAKE3).
  \item Chunks are concatenated in order to reconstruct the original file.
  \item The file is served to the caller and cached for future reads.
  \item Unsync/dehydrate evicts cached content or creates a physical
        \code{.tc} stub, depending on the surface.
\end{enumerate}

% ─────────────────────────────────────────────────────────────────────────────
\section{Credential Chain}

\tcfsd{} discovers age identities in order of precedence:

\begin{enumerate}
  \item \code{\$CREDENTIALS\_DIRECTORY/age-identity} --- systemd
        \code{LoadCredentialEncrypted}
  \item \code{\$SOPS\_AGE\_KEY\_FILE} --- path to an age key file
  \item \code{\$SOPS\_AGE\_KEY} --- literal age key content
  \item \filepath{\textasciitilde/.config/sops/age/keys.txt} --- default fallback
\end{enumerate}

Once loaded, the age identity decrypts SOPS-encrypted credential YAML files,
providing S3 access keys to the OpenDAL operator. An mtime watcher
auto-reloads credentials when the file changes on disk.

% ─────────────────────────────────────────────────────────────────────────────
\section{Phase Roadmap}

\begin{longtable}{clp{9cm}}
  \toprule
  \textbf{Phase} & \textbf{Status} & \textbf{Scope} \\
  \midrule
  \endhead
  0 & Complete & Repo foundation, SOPS migration, Rust workspace stubs \\
  1 & Complete & Core daemon + secrets + gRPC \\
  2 & Complete & Sync engine + chunking + NATS \\
  3 & Complete & FUSE driver + clean VFS hydration + physical stubs \\
  4 & Complete & K8s backend + HPA + full OpenTofu deploy \\
  5 & Complete & Release pipeline + packaging + docs site \\
  6 & Complete & macOS FUSE-T + Windows CFAPI skeleton + GitLab mirror \\
  7 & Complete & Benchmarks, fleet ops, Darwin RFC, production hardening \\
  \bottomrule
\end{longtable}

% ─────────────────────────────────────────────────────────────────────────────
\section{Infrastructure}

\subsection{Local Network (Bare-Metal)}

\begin{longtable}{ll}
  \toprule
  \textbf{Role} & \textbf{Address} \\
  \midrule
  \endhead
  SeaweedFS master-1         & 192.168.101.249:9333 \\
  SeaweedFS master-2         & 192.168.101.184:9333 \\
  SeaweedFS master-3         & 192.168.101.248:9333 \\
  Volume server (Drobo 5C)   & 192.168.101.171:8080 \\
  Filer / S3 gateway         & 192.168.101.146:8333 \\
  \bottomrule
\end{longtable}

\subsection{Civo Kubernetes}

\begin{longtable}{ll}
  \toprule
  \textbf{Property} & \textbf{Value} \\
  \midrule
  \endhead
  Cluster          & bitter-darkness-16657317 \\
  Namespace        & tcfs \\
  In-cluster SeaweedFS & seaweedfs.tcfs.svc.cluster.local:8333 \\
  In-cluster NATS  & nats.tcfs.svc.cluster.local:4222 \\
  Container image  & ghcr.io/jesssullivan/tcfsd \\
  \bottomrule
\end{longtable}

\end{document}
